Analyze

### 2023-10-06 analyze.qmd
Fix #8
Cross reference sections (#11)
WIP

---
title: Analyze
---

@@

library(scales)
library(cmdstanr)
library(posterior)
library(brms)
library(ggstance)
library(ggdist)
library(patchwork)
library(tidyverse)
source("functions.R")
```

```{r}
#| label: options
#| cache: false

dir.create("models", FALSE)
N_CORES <- as.numeric(Sys.getenv("N_CORES", unset = 1))
N_THREADS <- as.numeric(Sys.getenv("N_THREADS", unset = 2))
N_ITER <- as.numeric(Sys.getenv("N_ITER", unset = 2000))
N_SUBSET <- as.numeric(Sys.getenv("N_SUBSET"))
options(
  mc.cores = N_CORES,
  brms.backend = "cmdstanr"
)
theme_set(
  theme_linedraw(base_size = 10) +
    theme(
      strip.background = element_rect(fill = NA, color = NA),
      strip.text = element_text(colour = "black", hjust = 0),
      axis.text = element_text(size = rel(0.75)),
      panel.grid = element_blank()
    )
)
```

```{r}
dat <- readRDS("data/data.rds")
s <- readRDS("data/s.rds")
sraw <- readRDS("data/sraw.rds")
if (!is.na(N_SUBSET)) {
  pid_subset <- sample(unique(dat$pid), N_SUBSET)
  dat <- filter(dat, pid %in% pid_subset)
}
```

# Methods

@@

a <- quantile(demo$age, probs = c(.5, .1, .9), na.rm = TRUE)
a <- str_glue("{a[1]} ({a[2]}, {a[3]})")
g <- count(demo, gender, sort = TRUE) |>
  mutate(p = n / sum(n)) |>
  mutate(x = str_glue("{gender} ({number2(n)}, {percent2(p)})"))
n_countries <- length(unique(demo$country[!is.na(demo$country)]))
countries <- count(demo, country, sort = TRUE) |>
  mutate(p = n / sum(n)) |>
  mutate(x = str_glue("{country} ({number2(n)}, {percent2(p)})"))
```

The PowerWash Simulator case study database contains `r sraw$n_res_total` psychological item responses from `r sraw$n_sub` individuals. In this study, we only studied subjective well-being (mood), so did not consider the other five types of item responses in the database (enjoyment, focus, autonomy, competence, and immersion). Therefore, here we used data from `r s$n_sub` participants who provided `r s$n_obs` mood responses (`r s$n_obs_missing` with missing values, which we ignored) from `r s$n_ses` gameplay sessions.

These participants median age (1st & 9th deciles) was `r a`, and four most frequent responses to gender were `r g$x[1:4]`. They hailed from `r n_countries`, with four most represented countries being `r countries$x[1:4]`.

The study procedures were granted ethical approval by Oxford University’s Central University Research Ethics Committee (SSH_OII_CIA_21_011). All participants provided informed consent to participate and reported being 18 years or older.

## Data analysis

```{r}
#| label: data-wrangle-for-analysis

dat <- dat |>
  mutate(
    post = factor(
      row_number() > 1,
      levels = c(FALSE, TRUE),
      labels = c("0", "1")
    ),
    .by = c(pid, session)
  ) |>
  mutate(ps = interaction(pid, session, sep = "_")) |>
  mutate(
    cl = case_when(
      mood == 0 ~ "left",
      mood == 1 ~ "right",
      TRUE ~ "none"
    )
  )
```

@@

More formally, we modelled each mood response *i* in 1 to `r s$n_obs` from participant *j* in 1 to `r s$n_sub` and participant-session *k* in 1 to `r s$n_ses` with a normal distribution censored at the VAS limits [0, 1]. We then modelled the distribution's mean ($\eta$) on a population-level intercept ($\beta_0$) and contrast-coded measurement timepoint ($\beta_1$), their random per-player deviations $\alpha$, and a random intercept for player-sessions $\gamma_0$:

\begin{align*}
\text{mood}_{ijk} &\sim \text{CensNormal}(\eta_{ijk}, \sigma^2, 0, 1), \\
\eta_{ijk} &= \alpha_0 + \beta_{0j} +
  (\alpha_1 + \beta_{1j})\text{post}_{ijk} +
  \gamma_{0k}, \\
\begin{bmatrix} \alpha_0 \\ \alpha_1 \end{bmatrix} &\sim 
  \text{MVN}\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
  \begin{bmatrix} 
    \tau_{\alpha_0} &\rho_{{\alpha_0}{\alpha_1}} \\ 
    \rho_{{\alpha_0}{\alpha_1}} &\tau_{\alpha_1}
  \end{bmatrix}\right), \\
  \gamma_0 &\sim \text{Normal}(0, \tau_{\gamma_0}).
\end{align*}

```{r}
#| label: model-estimate
#| cache: false

contrasts(dat$post) <- c(-0.5, 0.5)

model <- bf(mood | cens(cl) ~ 1 + post + (1 + post | pid) + (1 | ps)) +
  gaussian()

fit <- brm(
  model,
  data = dat,
  silent = 0,
  threads = N_THREADS,
  iter = N_ITER,
  control = list(adapt_delta = .95),
  file = "models/brm-pre-post-pid-ps-censored"
)
```

```{r}
#| label: model-convergence
#| eval: false

as_draws(fit, variable = c("b_", "sd_", "cor_"), regex = TRUE) |> 
  summarise_draws(rhat)
pairs(fit$fit, pars = variables(fit)[1:7])
```

We analyzed the data with R [@rcoreteamLanguageEnvironmentStatistical2023] and used the brms package to estimate, via Stan's HMC sampling procedures, and post-process the models [@burknerBrmsPackageBayesian2017; @standevelopmentteamStanModelingLanguage2022]. We drew `r number2(ndraws(fit))` samples from the posterior distribution using brms' default prior distributions on all parameters, and used numerical and graphical checks to ensure model convergence and adequacy.

# Results

```{r}
#| label: fig-descriptives
#| include: false
#| fig-height: 2.2
#| fig-width: 7
#| fig-cap: '**A**. Histogram of session durations (note log10 x-axis). **B**. Summary of how many sessions each participant completed. **C**. Histogram of all mood ratings.'

p0 <- dat |>
  ggplot() +
  scale_x_continuous(
    expand = expansion(c(0.01, 0.01)),
  ) +
  scale_y_continuous(
    expand = expansion(c(0.001, 0.1)),
  ) +
  geom_histogram(
    col = "white",
    linewidth = .25
  )

dat_session_durations <- dat |>
  filter(duration == max(duration), .by = c(pid, session))

p_durations <- p0 %+%
  dat_session_durations +
  aes(duration) +
  scale_x_log10() +
  coord_cartesian(xlim = c(5, 900)) +
  labs(x = "Session duration (minutes)", y = "Sessions")

p_sessions <- p0 %+%
  dat_session_counts +
  coord_cartesian(xlim = c(0, 15)) +
  aes(sessions) +
  labs(x = "Sessions", y = "Participants")

(p_durations | p_sessions | p_mood) +
  plot_annotation(tag_level = "A")
```

```{r}
tmp <- list(
  dat_session_durations$duration,
  dat_session_counts$sessions,
  dat$mood,
  dat$mood[dat$post == 0],
  dat$mood[dat$post == 1]
) |> 
  map(~median_qi(.x, .width = 0.8, na.rm = TRUE)) |> 
  map(~str_glue_data(.x, "{y} [{ymin}, {ymax}]"))
```

The median session duration was `r tmp[[1]]` minutes [10 and 90 percentiles]; the median player contributed data from `r tmp[[2]]` sessions, and the median mood was `r tmp[[3]]` (pre-session: `r tmp[[4]]`, during play: `r tmp[[5]]`).

```{r}
#| label: fig-data
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**A**. Representative data from three participants (rows) showing scatterplots of mood responses over that session's (columns) duration. Each session's first mood response is highlighted in red, subsequent reports are blue. **B**. Horizontal histograms of all sessions' mood response for that person, with means and bootstrapped 95%CIs. **C**. Histograms of session-mean (**D**. person-mean) moods before (top) and during (bottom) sessions. **E**. Differences in session-mean (**F**. player-mean) mood differences (during session - pre-play). **G**. Scatterplot of person-mean mood reports at beginning (x-axis) and during game play sessions (y-axis)."

axislabels <- c(0, 0.5, 1.0)

# Plot: Example raw sessions
set.seed(99)
dat_example_sessions <- dat |>
  add_count(pid, session, name = "obs_per_session") |>
  filter(obs_per_session >= 4) |>
  mutate(session = as.numeric(as.factor(session)), .by = pid) |>
  mutate(session_per_person = length(unique(session)), .by = pid)
# default 
min_sessions <- 4 
n_filtered <- dat_example_sessions |> 
  distinct(pid, session_per_person) |> 
  filter(session_per_person >= 4) |> 
  nrow()
# pick anyone if there's not 3 pids with >= 4 sessions
if(n_filtered < 3) min_sessions <- 0
dat_example_sessions <- dat_example_sessions |>
 filter(session_per_person >= min_sessions) |>
  arrange(pid, session) |>
  filter(pid %in% sample(unique(pid), 3)) |>
  mutate(Session = str_glue("Session {session}"), Person = str_glue("Person {fct_anon(pid)}"))

p_mood_example_sessions <- dat_example_sessions |> 
  ggplot() +
  aes(duration / 60, mood, col = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_continuous(
    "Session duration (hours)"
  ) +
  scale_y_continuous(
    "Mood",
    limits = c(0, 1),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_point(size = 1.5, alpha = 1) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session),
    scales = "free_x"
  ) +
  theme(
    legend.position = "none",
    strip.text.y = element_blank()
  )

p_mood_example_all <- dat_example_sessions |> 
  mutate(Session = "All sessions") |> 
  ggplot() +
  aes(y = mood, x = post, col = post, fill = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_discrete(
    "Session",
    breaks = c(0, 1),
    labels = c("Pre", "During"),
    expand = expansion(c(0.2, 2.4))
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = axislabels,
    labels = axislabels
  ) +
  stat_summary(
    fun.data = mean_cl_boot,
    fatten = 2
  ) +
  geom_dots(side = "right", alpha = .33, binwidth = .075) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session)
  ) +
  theme(
    axis.text.y = element_blank(), 
    axis.title.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    axis.text.x = element_text(hjust = 0),
    strip.background.y = element_blank(),
    legend.position = "none"
  )

p_mood_example <- (p_mood_example_sessions | p_mood_example_all) +
  plot_layout(widths = c(92, 8))

# Plot: Raw mood histograms at pre and during play
p_mood_prepost_raw <- dat |> 
  mutate(post = factor(post, labels =  c("Pre-play", "During play"))) |>
  ggplot(aes(mood, fill = post)) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  geom_histogram(bins = 30, col = "white") +
  scale_y_continuous(
    "Observations",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  ) +
  scale_x_continuous(
    "Mood",
    expand = expansion(c(0.001)),
    breaks = axislabels
  ) +
  facet_wrap("post", ncol = 1, scales = "free_y") +
  theme(legend.position = "none")

# Plot: Difference histogram (sessions)
p_mood_difference_sessions <- p_mood_prepost_sessions$data |> 
  pivot_wider(names_from = post, values_from = mood) |> 
  ggplot(aes(`During play` - `Pre-play`)) +
  geom_histogram(bins = 50, col = "white") +
  geom_vline(xintercept = 0, linewidth = .33, col = "dodgerblue") +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    "Difference",
    expand = expansion(c(0.001))
  ) +
  coord_cartesian(xlim = c(-0.4, 0.6))

# Plot: Difference histogram (players)
p_mood_difference_players <- p_mood_difference_sessions %+% 
  summarise(
    p_mood_difference_sessions$data,
    across(`Pre-play`:`During play`, ~mean(.x, na.rm = TRUE)),
    .by = pid
  ) +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0.001, 0.1))
  )

p_mood_biscatter_sessions <- p_mood_difference_sessions$data |> 
  ggplot(aes(`Pre-play`, `During play`)) +
  scale_x_continuous(
    expand = expansion(0.01),
    breaks = axislabels,
    labels = axislabels
  ) +
  scale_y_continuous(
    expand = expansion(0.01),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_point(
    alpha = .25, size = 1, shape = 1
  ) +
  geom_abline(linewidth = .33, col = "dodgerblue") +
  theme(aspect.ratio = 1)

p_mood_biscatter_players <- p_mood_biscatter_sessions %+% 
  p_mood_difference_players$data

p_mood_example / 
  (
    p_mood_prepost_sessions | 
      p_mood_prepost_players | 
      p_mood_difference | 
      p_mood_biscatter_players
  ) +
  plot_layout(heights = c(4.5, 5.5)) +
  plot_annotation(tag_levels = "A")
```

We then turned to our primary research question: To what extent does player mood in PWS change from pre-play to during play? First, we visualized the relevant data in @fig-data: Panel A shows mood responses from three example participants' first five sessions of play. @fig-data B then shows histograms of those data to facilitate visual comparison of the raw data. We then calculated sessions' and participants' mean moods at the beginning and during each session (@fig-data C, D, and G), and their differences (E and F). Visual inspection suggested greater mood during play compared to the pre-play measures at both the level of sessions and players.

@@

h_avg_ <- hypothesis(fit, hs)
h_avg <- tibble(h_avg_$samples)
names(h_avg) <- names(hs)
h_avg |> 
  summarise_draws(
    mean = ~number(mean(.), .001), 
    ~number2(quantile2(., probs = c(.025, .975)), .001), 
    pd = ~percent2(Pr(. > 0))
  ) |> 
  mutate(CI = str_glue("[{q2.5}, {q97.5}]")) |> 
  select(` ` = variable, Estimate = mean, CI, pd) |> 
  knitr::kable()
```

The model results confirmed the visual impressions described above: @tbl-avg indicates, with great certainty, that mood increased by about [`r paste0(map(h_avg_$hypothesis[3,4:5], ~number(.x, .001)), collapse = ", ")`] units (@fig-coef, "Raw").

@@

posterior_pop <- as_draws_df(
  fit, 
  c("sd_pid__Intercept", "sd_pid__post1", "sigma", "b_post1")
) |> 
  mutate(
    total_residual = sd_pid__Intercept + sd_pid__post1 + sigma,
    b_post1_scaled = b_post1 / total_residual
  ) |> 
  tibble() |> 
  select(-starts_with("."))

posterior_pop_sum <- posterior_pop |> 
  summarise_draws(
    mean = ~number(mean(.), .001), 
    ~number2(quantile2(., probs = c(.025, .975)), .001)
  ) |> 
  mutate(
    x = str_glue("{mean} [{q2.5}, {q97.5}]")
  )
```

To further interpret the magnitude of the difference in moods between pre- and during play, we compared the contrast to the natural variations in mood as estimated by the model (standard deviation of the between-person intercept heterogeneity ($\tau_{\alpha_0}$ = `r posterior_pop_sum$x[1]`), between-person pre-during difference heterogeneity ($\tau_{\alpha_1}$ = `r posterior_pop_sum$x[2]`) and to the residual standard deviation ($\sigma$ = `r posterior_pop_sum$x[3]`). The standardized (with the total residual variation from the three above components) pre-play -- during play contrast was `r posterior_pop_sum$x[6]` (@fig-coef, "Scaled"), a not ignorable difference.

@@

posterior_pid <- coef(fit, summary = FALSE)$pid[,,"post1"]
posterior_pid_scaled <- apply(
  posterior_pid, 2, 
  function(x) x / posterior_pop$total_residual
) |> 
  apply(2, function(x) mean_qi(x)) |> 
  bind_rows(.id = "pid")
posterior_pid <- posterior_pid |> 
  apply(2, function(x) mean_qi(x)) |> 
  bind_rows(.id = "pid")

posterior_pid_sum <- bind_rows(
  "b_post1" = posterior_pid,
  "b_post1_scaled" = posterior_pid_scaled,
  .id = "name"
)

x <- posterior_pop |> 
  mutate(
    prop_pos = pnorm(0, b_post1, sd_pid__post1, lower.tail = FALSE)
  ) |> 
  mean_qi(prop_pos) |> 
  mutate(x = str_glue("{percent(prop_pos, .1)} [{number(.lower*100, .1)}, {number(.upper*100, .1)}]")) |> 
  pull(x)
```

We then turned our attention to between-person differences in the difference in during and pre-play mood. With a sample of `r s$n_sub`, it is not surprising that the population-level contrast is credibly different from zero. We therefore asked the subsequent question: What proportion of the population from which the sample was drawn would we expect there to be an elevation in mood during play? The boxplots in @fig-coef show the distribution of player-specific contrasts estimated by the model, which indicated that `r x` of players in the population experience a positive shift in mood during play, compared to their moods during login.

```{r}
#| label: fig-coef
#| include: true
#| fig.width: 3
#| fig.height: 1.4
#| fig.cap: "Difference in mood during play compared to mood at session start. Points and intervals indicate posterior means and 95%CIs of the population level difference; boxplots summarize the player-level estimates. Raw estimates are on the VAS scale, whereas Scaled estimates indicate contrasts divided by the total residual deviance of the model."

posterior_pop |> 
  select(starts_with("b_")) |> 
  pivot_longer(everything()) |>
  group_by(name) |> 
  mean_qi(value, .width = 0.95) |> 
  ggplot(aes(value, name)) +
  scale_x_continuous(
    "During play - pre-play contrast",
    breaks = extended_breaks(7)
  ) +
  scale_y_discrete(
    breaks = c("b_post1", "b_post1_scaled"),
    labels = c("Raw", "Scaled")
  ) +
  coord_cartesian(xlim = c(-0.25, 0.65)) +
  geom_vline(xintercept = 0, linewidth = .33, col = "dodgerblue") +
  geom_boxploth(
    data = posterior_pid_sum,
    aes(x = y, group = name), 
    outlier.shape = NA,
    notch = TRUE,
    size = 0.25,
    width = .2
  ) +
  geom_pointrangeh(
    aes(xmin = .lower, xmax = .upper),
    fatten = 2.5
  ) +
  theme(
    axis.title.y = element_blank()
  )
```

# Bibliography

::: {#refs}
:::

>>>>

# Analyze {#sec-analyze}

@@

library(scales)
library(cmdstanr)
library(posterior)
library(brms)
library(ggstance)
library(ggdist)
library(patchwork)
library(tidyverse)
source("R/functions.R")
source("R/fit_cached.R")
source("R/common.R")
dir.create("models", FALSE)
```

```{r}
dat <- readRDS("data/data.rds")

if (!is.na(N_SUBSET)) {
  pid_subset <- sample(unique(dat$pid), N_SUBSET)
  dat <- filter(dat, pid %in% pid_subset)
}
```

@@

a <- quantile(demo$age, probs = c(.5, .1, .9), na.rm = TRUE)
a <- str_glue("{a[1]} ({a[2]}, {a[3]})")
```

- N participants: `r number2(length(unique(dat$pid)))`
- N sessions: `r number2(length(unique(dat$ps)))`
- age: `r a`
- gender:

```{r}
#| label: demographics-gender

count(demo, gender, sort = TRUE) |>
  mutate(p = n / sum(n)) |>
  mutate(x = str_glue("{gender} ({number2(n)}, {percent2(p)})"))
```

- countries:
```{r}
#| label: demographics-countries

n_countries <- length(unique(demo$country[!is.na(demo$country)]))
count(demo, country, sort = TRUE) |>
  mutate(p = n / sum(n)) |>
  mutate(x = str_glue("{country} ({number2(n)}, {percent2(p)})"))
```

## Descriptives

```{r}
#| label: fig-descriptives
#| fig-height: 2.2
#| fig-width: 7
#| fig-cap: '**A**. Histogram of session durations (note log10 x-axis). **B**. Summary of how many sessions each participant completed. **C**. Histogram of all mood ratings.'

p0 <- dat |>
  ggplot() +
  scale_x_continuous(
    expand = expansion(c(0.01, 0.01)),
  ) +
  scale_y_continuous(
    expand = expansion(c(0.001, 0.05)),
  ) +
  geom_histogram(
    col = "white",
    bins = 50,
    linewidth = .25,
    boundary = 0
  )

dat_session_durations <- dat |>
  filter(hours == max(hours), .by = c(pid, session))

p_durations <- p0 %+%
  dat_session_durations +
  aes(hours) +
  labs(x = "Session duration (hours)", y = "Sessions")

p_sessions <- p0 %+%
  dat_session_counts +
  coord_cartesian(xlim = c(0, 75)) +
  aes(sessions) +
  labs(x = "Sessions", y = "Participants")

(p_durations | p_sessions | p_mood) +
  plot_annotation(tag_levels = "A")
```

```{r}
tmp <- list(
  dat_session_durations$hours,
  dat_session_counts$sessions,
  dat$mood,
  dat$mood[dat$post == 0],
  dat$mood[dat$post == 1]
) |> 
  map(~median_qi(.x, .width = 0.8, na.rm = TRUE)) |> 
  map(~mutate(.x, across(c(y, ymin, ymax), ~number2(., .01))) |> str_glue_data("{y} [{ymin}, {ymax}]"))
```

The median session duration was `r tmp[[1]]` hours [10 and 90 percentiles]; the median player contributed data from `r tmp[[2]]` sessions, and the median mood was `r tmp[[3]]` (pre-session: `r tmp[[4]]`, during play: `r tmp[[5]]`).

```{r}
#| label: fig-data
#| include: true
#| fig-width: 8
#| fig-height: 4.6
#| fig-cap: "**A**. Representative data from three participants (rows) showing scatterplots of mood responses over that session's (columns) duration. Each session's first mood response is highlighted in red, subsequent reports are blue. **B**. Horizontal histograms of all sessions' mood response for that person, with means and bootstrapped 95%CIs. **C**. Histograms of session-mean (**D**. person-mean) moods before (top) and during (bottom) sessions. **E**. Differences in session-mean (**F**. player-mean) mood differences (during session - pre-play). **G**. Scatterplot of person-mean mood reports at beginning (x-axis) and during game play sessions (y-axis)."

# Plot: Example raw sessions
set.seed(99)
dat_example_sessions <- dat |>
  add_count(pid, session, name = "obs_per_session") |>
  filter(obs_per_session >= 4) |>
  mutate(session = as.numeric(as.factor(session)), .by = pid) |>
  mutate(session_per_person = length(unique(session)), .by = pid)

# default 
min_sessions <- 8
n_filtered <- dat_example_sessions |> 
  distinct(pid, session_per_person) |> 
  filter(session_per_person >= 8) |> 
  nrow()

# pick anyone if there's not 3 pids with >= 4 sessions
if(n_filtered < 3) min_sessions <- 0
dat_example_sessions <- dat_example_sessions |>
  filter(session_per_person >= min_sessions) |>
  arrange(pid, session) |>
  filter(pid %in% sample(unique(pid), 3)) |>
  mutate(Session = str_glue("Session {session}"), Person = str_glue("Person {fct_anon(pid)}"))

p_mood_example <- dat_example_sessions |> 
  filter(session <= 8) |> 
  ggplot() +
  aes(hours, mood, col = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_continuous(
    "Session duration (hours)"
  ) +
  scale_y_continuous(
    "Mood",
    limits = c(0, 1),
  ) +
  geom_point(size = 1.5, alpha = 1) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session),
    scales = "fixed"
  ) +
  theme(
    legend.position = "none",
    strip.text.y = element_blank()
  )

# Plot: Raw mood histograms at pre and during play
p_mood_prepost_raw <- dat |> 
  mutate(post = factor(post, labels =  c("Pre-play", "During play"))) |>
  ggplot(aes(mood, fill = post)) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  geom_histogram(bins = 30, col = "white") +
  scale_y_continuous(
    "Observations",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  ) +
  scale_x_continuous(
    "Mood",
    expand = expansion(c(0.001))
  ) +
  facet_wrap("post", ncol = 1, scales = "free_y") +
  theme(legend.position = "none")

# Plot: Difference histogram (sessions)
p_mood_difference_sessions <- p_mood_prepost_sessions$data |> 
  pivot_wider(names_from = post, values_from = mood) |> 
  mutate(Difference = `During play` - `Pre-play`) |> 
  drop_na(Difference) |> 
  ggplot(aes(Difference)) +
  geom_histogram(bins = 50, col = "white") +
  geom_vline(xintercept = 0, linewidth = .33, col = "dodgerblue") +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    expand = expansion(c(0.001))
  ) +
  coord_cartesian(xlim = c(-.4, .6))

# Plot: Difference histogram (players)
p_mood_difference_players <- p_mood_prepost_players$data |> 
  pivot_wider(names_from = post, values_from = mood) |> 
  mutate(Difference = `During play` - `Pre-play`) |> 
  drop_na(Difference) |> 
  ggplot(aes(Difference)) +
  geom_histogram(bins = 50, col = "white") +
  geom_vline(xintercept = 0, linewidth = .33, col = "dodgerblue") +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    expand = expansion(c(0.001))
  ) +
  coord_cartesian(xlim = c(-.4, .6))

p_mood_biscatter_sessions <- p_mood_difference_sessions$data |> 
  ggplot(aes(`Pre-play`, `During play`)) +
  scale_x_continuous(
    expand = expansion(0.01)
  ) +
  scale_y_continuous(
    expand = expansion(0.01)
  ) +
  geom_point(
    alpha = .25, size = 1, shape = 1
  ) +
  geom_abline(linewidth = .33, col = "dodgerblue") +
  theme(aspect.ratio = 1)

p_mood_biscatter_players <- p_mood_difference_players$data |> 
  ggplot(aes(`Pre-play`, `During play`)) +
  scale_x_continuous(
    expand = expansion(0.01)
  ) +
  scale_y_continuous(
    expand = expansion(0.01)
  ) +
  geom_point(
    alpha = .25, size = 1, shape = 1
  ) +
  geom_abline(linewidth = .33, col = "dodgerblue") +
  theme(aspect.ratio = 1)

p_mood_example / 
  (
    p_mood_prepost_sessions | 
      p_mood_prepost_players | 
      p_mood_difference | 
      p_mood_biscatter_players
  ) +
  plot_layout(heights = c(5, 5)) +
  plot_annotation(tag_levels = "A")
```

## Model

@@

More formally, we modelled each mood response *i* in 1 to `r number2(nrow(dat))` from participant *j* in 1 to `r number2(length(unique(dat$pid)))` and participant-session *k* in 1 to `r number2(length(unique(dat$ps)))` with a normal distribution censored at the VAS limits [0, 1]. We then modelled the distribution's mean ($\eta$) on a population-level intercept ($\beta_0$) and contrast-coded measurement timepoint ($\beta_1$), their random player ($\alpha$) and player-session ($\gamma$) deviations,

$$
\begin{align*}
\text{mood}_{ijk} &\sim \text{CensNormal}(\eta_{ijk}, \sigma^2, 0, 1), \\
\eta_{ijk} &= \alpha_0 + \beta_{0j} + \gamma_{0k} +
  (\alpha_1 + \beta_{1j})\text{post}_{ijk}, \\
\begin{bmatrix} \alpha_0 \\ \alpha_1 \end{bmatrix} &\sim 
  \text{MVN}\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
  \begin{bmatrix} 
    \tau_{\alpha_0} &\rho_{{\alpha_0}{\alpha_1}} \\ 
    \rho_{{\alpha_0}{\alpha_1}} &\tau_{\alpha_1}
  \end{bmatrix}\right), \\
\begin{bmatrix} \gamma_0 \end{bmatrix} &\sim 
  \text{Normal}\left(\begin{bmatrix} 0 \end{bmatrix}, 
  \begin{bmatrix} 
    \tau_{\gamma_0}
  \end{bmatrix}\right).
\end{align*}
$$

```{r}
#| label: model-estimate-brms
#| cache: false

model <- bf(
  mood | cens(cl) ~ 
    1 + post + 
    (1 + post | pid) + 
    (1 | ps)
) +
  gaussian()

fit <- brm(
  model,
  data = dat,
  silent = 0,
  iter = N_ITER,
  control = list(adapt_delta = .95),
  file = "models/brm-prepost-pid-ps-censored"
)
```

```{r}
#| label: fig-convergence
#| fig-cap: Convergence diagnostic plot showing bivariate scatterplots of the model's population-level parameters' posterior draws.

pairs(
  fit, 
  variable = c("b_", "sd_", "cor_"), 
  regex = TRUE,
  off_diag_args = list(size = .33, shape = 1)
  )
```

## Population level effects

@@

h_avg_ <- hypothesis(fit, hs)
h_avg <- tibble(h_avg_$samples)
names(h_avg) <- names(hs)
h_avg_tab <- h_avg |> 
  summarise_draws(
    mean = ~number(mean(.), .001), 
    ~number2(quantile2(., probs = c(.025, .975)), .001), 
    pd = ~percent2(Pr(. > 0))
  ) |> 
  mutate(CI = str_glue("[{q2.5}, {q97.5}]")) |> 
  select(` ` = variable, Estimate = mean, CI, pd)
h_avg_tab |>   
  knitr::kable()
```

@@

posterior_pop <- as_draws_df(
  fit, 
  c("sd_", "sigma", "b_post1"),
  regex = TRUE
) |> 
  mutate(
    tr = sd_pid__Intercept + sd_pid__post1 + sd_ps__Intercept + sigma,
    b_post1_z = b_post1 / tr,
    prop_pos = pnorm(0, b_post1, sd_pid__post1, lower.tail = FALSE)
  ) |> 
  tibble() |> 
  select(-starts_with("."))
posterior_pop |> 
  summarize_draws(
    mean, ~quantile2(., probs = c(0.025, 0.975))
  ) |> 
  knitr::kable(digits = 3)
```

## Heterogeneity

@@

posterior_pid <- coef(fit, summary = FALSE)$pid[,,"post1"]
posterior_pid_scaled <- apply(
  posterior_pid, 2, 
  function(x) x / posterior_pop$tr
) |> 
  apply(2, function(x) mean_qi(x)) |> 
  bind_rows(.id = "pid")
posterior_pid <- posterior_pid |> 
  apply(2, function(x) mean_qi(x)) |> 
  bind_rows(.id = "pid")

posterior_pid_sum <- bind_rows(
  "b_post1" = posterior_pid,
  "b_post1_z" = posterior_pid_scaled,
  .id = "name"
) |> tibble()
```

```{r}
#| label: fig-coef
#| include: true
#| fig.width: 4
#| fig.height: 2.6
#| fig.cap: Histograms of person-specific contrasts on the raw scaled and standardized scale.

posterior_pid_sum |> 
  ggplot(aes(y, name)) +
  stat_slab(
    density = "histogram", 
    breaks = 100, 
    slab_color = "white", 
    outline_bars = TRUE, 
    linewidth = .25
  ) +
  geom_vline(xintercept = 0, linewidth = .33, col = "dodgerblue") +
  scale_x_continuous(
    "During play - pre-play contrast",
    breaks = extended_breaks(7)
  ) +
  scale_y_discrete(
    breaks = c("b_post1", "b_post1_z"),
    labels = c("Raw", "Scaled"),
    expand = expansion(c(0.05, 0.01))
  ) +
  theme(
    axis.title.y = element_blank()
  ) +
  coord_cartesian(ylim = c(1, 2.5), xlim = c(-0.2, 0.4))
```

```{r}
p_mood_difference_players$data |> 
  left_join(posterior_pid) |> 
  left_join(count(dat, pid)) |> 
  ggplot(aes(Difference, y, alpha = n, size = n)) +
  scale_x_continuous(
    "Difference"
  ) +
  scale_y_continuous(
    "Modelled difference"
  ) +
  coord_cartesian(
    xlim = c(-1, 1), ylim = c(-.5, .5)
  ) +
  geom_point() +
  geom_abline() +
  theme(aspect.ratio = 1)
```
